An Exposition on the Propriety of Restricted Boltzmann Machines

### 2016-07-28 presentations/jsm2016/index.Rmd
added motivation and conclusion

<div class="double">
<p class="double-flow">
Restricted Boltzmann machine (RBM) with two layers - hidden ($\mathcal{H}$) and visible ($\mathcal{V}$) [@friedman2001elements].
</p>
<p class="double-flow">
<img src="images/rbm.png" width="100%"></br>
</p>
</div>
<img src="images/visibles_image.png" width="100%"></br>
Used for image classification. Each image pixel is a node in the visible layer. The output creates features, passed to supervised learning.

@@

@@

> 1. *Trick prior.* Cancel out the normalizing term, full conditionals normally dist'd. 
   $$
    \pi(\theta) \propto \gamma(\theta)^n \exp\left(-\frac{1}{2C_{1}} \theta_{main}' \theta_{main} -\frac{1}{2C_{2}} \theta_{int}' \theta_{int}\right),
   $$
    where $\gamma(\theta)$ is the normalizing term [@li2014biclustering].
> 2. *Truncated Normal prior.* Use two independent truncated spherical normal distributions as priors for $\theta_{main}$ and $\theta_{int}$ with $\sigma_{int} < \sigma_{main}$. Not conjugate $\Rightarrow$ geometric adaptive MH step [@zhou2014some] and calculation of $\gamma(\theta)$. 
> 3. *Marginalized likelihood* Marginalize out $h$ in $f_{\theta}(x)$, use trunc. Normal prior. 
    $$
    g_{ \theta}(v) = \sum\limits_{\in \{-1,1\}^H} \exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)
    $$

@@

```{r models, cache=TRUE}
source("../../4x4 model/functs_sample.R")

@@

all_statistics %>%
  gather(statistic, value, -iter, -method, -starts_with("v")) %>%
  filter(grepl("image", statistic) & !is.na(value)) %>%
  separate(statistic, into = c("junk", "statistic")) %>%
  mutate(statistic = factor(statistic, labels = paste("Image", 1:length(unique(statistic))))) %>%
  left_join(distn_good %>% select(-image_id)) %>% 
  ggplot() +
  geom_density(aes(value, colour = method, fill = method), alpha = .2) +
  geom_vline(aes(xintercept = prob)) +
  facet_wrap(~statistic, scales = "free_x") + 
  ylab("Posterior Density") + xlab("Probability of Image") +
  scale_colour_discrete("Method", labels = c("Marginalized likelihood", "Trick prior", "Truncated Normal prior")) +
  scale_fill_discrete("Method", labels = c("Marginalized likelihood", "Trick prior", "Truncated Normal prior")) -> p.models
```
```{r models-plot, fig.width=8, fig.height=4}
p.models
```
Each of these rigorous fitting methods seeks to merely replicate the empirical data distribution, which is the optimal nonparametric solution.

>>>>

<div class="double">
<p class="double-flow">
Restricted Boltzmann machine (RBM) with two layers - hidden ($\mathcal{H}$) and visible ($\mathcal{V}$) [smolensky1986information].
</p>
<p class="double-flow">
<img src="images/rbm.png" width="100%"></br>
</p>
</div>
<img src="images/visibles_image.png" width="100%"></br>
Used for image classification. Each image pixel is a node in the visible layer. The output creates features, passed to supervised learning.

@@

## Why do I care?

Current heuristic fitting methods seem to work for classification. Beyond classification, RBMs are generative models:

> To generate data from an RBM, we can start with a random state in one of the layers and then perform alternating Gibbs sampling.
[@hinton2006fast]

Can we fit a model that generates data that looks like data?

@@

> 1. *Trick prior.* Cancel out the normalizing term, full conditionals normally dist'd. 
   $$
    \pi(\theta) \propto \gamma(\theta)^n \exp\left(-\frac{1}{2C_{1}} \theta_{main}' \theta_{main} -\frac{1}{2C_{2}} \theta_{int}' \theta_{int}\right),
   $$
    where $\gamma(\theta)$ is the normalizing term [@li2014biclustering].
> 2. *Truncated Normal prior.* Use two independent truncated spherical normal distributions as priors for $\theta_{main}$ and $\theta_{int}$ with $\sigma_{int} < \sigma_{main}$. Not conjugate $\Rightarrow$ geometric adaptive MH step [@zhou2014some] and calculation of $\gamma(\theta)$. 
> 3. *Marginalized likelihood* Marginalize out $h$ in $f_{\theta}(x)$, use trunc. Normal prior. 
    $$
    g_{ \theta}(v) = \sum\limits_{h \in \{-1,1\}^H} \exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)
    $$

@@

```{r models, cache=TRUE}

##TODO: Add empirical line

source("../../4x4 model/functs_sample.R")

@@

all_statistics %>%
  gather(statistic, value, -iter, -method, -starts_with("v")) %>%
  filter(grepl("image", statistic) & !is.na(value)) %>%
  separate(statistic, into = c("junk", "statistic")) %>%
  mutate(statistic = factor(statistic, labels = paste("Image", 1:length(unique(statistic))))) %>%
  left_join(distn_good %>% select(-image_id)) %>% 
  ggplot() +
  geom_density(aes(value, y=..scaled.., colour = method, fill = method), alpha = .2) +
  geom_vline(aes(xintercept = prob)) +
  facet_wrap(~statistic, scales = "free_x") + 
  ylab("Scaled Posterior Density") + xlab("Probability of Image") +
  scale_colour_discrete("Method", labels = c("Marginalized likelihood", "Trick prior", "Truncated Normal prior")) +
  scale_fill_discrete("Method", labels = c("Marginalized likelihood", "Trick prior", "Truncated Normal prior")) -> p.models
```
```{r models-plot, fig.width=8, fig.height=4}
p.models
```
Each of these rigorous fitting methods seeks to merely replicate the empirical data distribution, which is the likelihood maximizer.

## Wrapping up

- While RMBs can be useful for classification, in the context of a statistical model as a representation of data, RBMs poor fit due to
    1. *near-degeneracy*, 
    2. *instability*, and
    3. *uninterpretability*.
- Rigorous fitting methodology is possible but slow, and ultimately results in the empirical data distribution.
